Revisiting~the~Classics: Online~RL~in~the~Programmable~Dataplane

### 2021-10-13 paper/noms-short-paper.tex
One page shorter, made ugly as per IEEE reqs.
3 pages to go.

% !TeX document-id = {5530719d-34df-4dd8-b4b5-e6ed092c2b36}
% !TeX program = pdflatex
% !BIB program = biber
\documentclass[
sigconf,natbib=false
,anonymous=true
,10pt
]{acmart}

\usepackage[subtle,bibnotes,charwidths,mathdisplays,indent,lists]{savetrees}

@@

\usepackage[%
%false
]{anonymous-acm}

@@

\usepackage[title]{appendix}

@@

\usepackage[font=footnotesize]{subcaption}
\usepackage{awesomebox}

@@

\usepackage{url}
\usepackage{hyperref}
\usepackage[nameinlink]{cleveref}
\newcommand{\crefrangeconjunction}{--}
\crefname{table}{table}{tables}

@@

@@

% Copyright
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference info
%\setcopyright{none}
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

%\settopmatter{printacmref=false, printccs=false, printfolios=true}
%\settopmatter{printfolios=true}
\settopmatter{printacmref=false, printccs=true, printfolios=true}

% DOI
\acmDOI{12345}

% ISBN
\acmISBN{67890}

%Conference
\acmConference[SOSR '21]{Symposium of SDx Research 2021}{September \numrange{20}{21}, 2021}{Online}
\acmYear{2021}
%\copyrightyear{}

%% {} with no args suppresses printing of the price
%\acmPrice{15.00}
\acmPrice{}

@@

\newcounter{insightc}
\newenvironment{insight}
	{
		\begin{tipblock}\refstepcounter{insightc}\textbf{Insight \theinsightc:}\em
	}
	{
		\end{tipblock}
	}

@@

%	\author{Paper \# XXX, XXX pages}
	 \author{Kyle A. Simpson}
	 \orcid{0000-0001-8068-9909}
	 \affiliation{%
	   \institution{University of Glasgow}
	   \city{Glasgow} 
	   \country{Scotland}
	 }
	 \email{k.simpson.1@research.gla.ac.uk}
	 \author{Dimitrios P. Pezaros}
\orcid{0000-0003-0939-378X}
	 \affiliation{%
	\institution{University of Glasgow}
	\city{Glasgow} 
	\country{Scotland}
}
\email{Dimitrios.Pezaros@gla.ac.uk}

\setlength{\abovecaptionskip}{5pt}
\setlength{\belowcaptionskip}{0pt}

\setlength{\textfloatsep}{3.0pt plus 0.0pt minus 3.0pt}
\setlength{\floatsep}{3.0pt plus 0.0pt minus 2.0pt}

\setlength{\dbltextfloatsep}{3.0pt plus 0.0pt minus 3.0pt}
\setlength{\dblfloatsep}{3.0pt plus 0.0pt minus 2.0pt}

@@

We present \emph{\approachshort{}}---\approach{}---the first work to bring \emph{online reinforcement learning} to the dataplane.
\approachshort{} makes online learning possible in constrained SmartNIC hardware by returning to classical RL techniques which simplify update logic and enable parallel processing.
Compared to offloading, we achieve a \SI{21}{$\times$} reduction in \num{99.99}\nthscript{th} tail inference times to \SI{34}{\micro\second}, and \SI{9.9}{$\times$} improvement in online throughput for real-world policy designs.
In-NIC execution eliminates PCIe transfers, and our asynchronous compute model ensures minimal impact on traffic carried by a co-hosted P4 dataplane.
\approachshort{}'s design scales with additional resources at compile-time to improve upon both decision latency and throughput, improving on offloading, and are quickly reconfigurable at runtime compared to reinstalling device firmware.

@@

\begin{CCSXML}
	<ccs2012>
	<concept>
	<concept_id>10003033.10003068.10003069</concept_id>
	<concept_desc>Networks~Data path algorithms</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10003033.10003099.10003102</concept_id>
	<concept_desc>Networks~Programmable networks</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10003752.10010070.10010071.10010261</concept_id>
	<concept_desc>Theory of computation~Reinforcement learning</concept_desc>
	<concept_significance>300</concept_significance>
	</concept>
	<concept>
	<concept_id>10010147.10010257.10010258.10010261</concept_id>
	<concept_desc>Computing methodologies~Reinforcement learning</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Networks~Data path algorithms}
\ccsdesc[500]{Networks~Programmable networks}
\ccsdesc[500]{Computing methodologies~Reinforcement learning}

@@

\setlength{\aweboxleftmargin}{0.12\linewidth}
\setlength{\aweboxcontentwidth}{1.97\linewidth}

@@

%The question we investigate is: can online RL be brought to the dataplane by returning to these computationally simpler methods, to act on locally extracted state?
%Can it be made more efficient \emph{by dataplane hardware}?
%Through this work, \approachshort{}, we can comfortably answer ``yes'' on both counts.
%Through \approachshort{}, w
We show that online RL can be brought to the dataplane by such methods, where it can act on locally extracted state and is made \emph{more efficient by dataplane hardware}.
In particular, we exploit how SmartNIC devices often expose general-purpose compute to provide path-adjacent, on-chip RL in the dataplane (\cref{fig:netro-arch}).
As many of these devices have engineering and development histories which predate P4, general compute beyond P4's limits~\parencite{p4-psa} is surprisingly common.
By executing on spare compute units, we prevent packet stalling and offer quick runtime reconfigurability.
%We look backwards to classical RL techniques (tile-coded policies, single-step RL algorithms), alongside the adaptations introduced by existing in-network ML (such as quantisation).
%Such adaptations unlock further improvements in parallel inference and update processing.
This paper contributes:
\begin{itemize}
	\item An analysis of why in-NIC RL is needed and best-placed to interact with the network, made feasible by classical RL methods and quantisation (\cref{sec:motivation}),
	\item \emph{\approachshort{}}: a general-purpose in-NIC RL agent which scales with allocated device resources to meet latency or throughput demands of network traffic analysis (\cref{sec:design}),
	\item \emph{ParSa}, a wait-free, parallel, online RL algorithm to accelerate tile coded policy inference and updates (\cref{alg:parsa}),
	\item In-depth evaluation of how \approachshort{} affects carried dataplane traffic, performs under different policy sizes (simple/complex state), and improves on explicit offloading with a \SI{15}{$\times$} latency reduction compared to commodity hardware (\SI{21}{$\times$} for 99.99\nthscript{th} tail latencies) and an order of magnitude improvement in online throughput (\cref{sec:evaluation}).
	\item A description of how \approachshort{} would integrate with state-of-the-art PDP applications to perform fully in-NIC, fast, automated DDoS mitigation (\cref{sec:potential-integrations}).
\end{itemize}

@@

\begin{figure}
	\begin{subfigure}{0.45\linewidth}
		\centering
		\resizebox{0.8\linewidth}{!}{
			\centering
			\begin{tikzpicture}
				\node[circle, draw] (state) {$S$};
				\node[circle, draw] (state') at ($(state) + (0, -2)$) {$S'$};

				\node (agent) at ($(2, 0) + (state)$) {Agent};
				\node[below of=agent] (action) {Action $A$};

				\node[right of=state'] {$+ R$};

				\draw[->] (state) -- (state') node[midway, right] {$A$};

				\draw[dotted, ->, bend left = 30] (state) -- (agent);
				\draw[->] (agent) -- (action);
				\draw[dotted, ->] (action) -- (state);
		\end{tikzpicture}}
		\caption{Theory: state measurement, action computation, and learning are zero-cost.}
	\end{subfigure}
	\hspace{0.05\linewidth}
	\begin{subfigure}{0.45\linewidth}
		\centering
		\resizebox{0.6\linewidth}{!}{
			\centering
			\begin{tikzpicture}
				\node[circle, draw] (state) {$S$};
				\node[circle, draw] (state') at ($(state) + (0, -1.5)$) {$S'$};
				\node[circle, draw] (state'') at ($(state') + (0, -1.5)$) {$S''$};

				\node (agent) at ($(2, 0) + (state)$) {Agent};
				\node[below of=agent] (action) {Action $A$};

				\node[right of=state''] {$+ R$};

				\draw[->] (state) -- (state') node[midway, right] {$\varnothing$};
				\draw[->] (state') -- (state'') node[midway, right] {$A$};

				\draw[dotted, ->, bend left = 30] (state) -- (agent) node[midway, above] {$t_1$};
				\draw[->] (agent) -- (action) node[midway, right] {$t_2$};
				\draw[dotted, ->] (action) -- (state') node[midway, below] {$t_3$};
		\end{tikzpicture}}
		\caption{Reality: policy costs lead to \emph{state drift}---over a delay $t_1+t_2+t_3$, state $S$ becomes $S'$.}
	\end{subfigure}
	\caption{Asynchronous RL delays and state slippage.\label{fig:state-slip}}
\end{figure}

@@

\begin{figure*}
	\centering
	\begin{subfigure}{0.47\linewidth}
		\centering
		\includegraphics[keepaspectratio, width=0.78\linewidth]{figures/ind}
		\caption{\Indfw{} (offline throughput-optimal). \emph{Workers} independently pull commands from (and push actions to) the environment, locking policy access for updates.\label{fig:single-and-parallel:single}}
	\end{subfigure}
	\hspace{0.04\linewidth}
	\begin{subfigure}{0.47\linewidth}
		\centering
		\includegraphics[keepaspectratio, width=0.8\linewidth]{figures/coop}
		\caption{\Coopfw{} (online-optimal). A single \emph{controller} delegates RL computation and updates to many \emph{minion} threads, who operate on independent subtasks.\label{fig:single-and-parallel:parallel}}
	\end{subfigure}
	\caption{\approachshort{}'s compute strategies scale to fit device capacity according to either latency or throughput needs.\label{fig:single-and-parallel}}
\end{figure*}

\Cref{fig:netro-arch,fig:single-and-parallel} outline our design on Netronome SmartNIC hardware in pursuit of this goal: unused device resources beyond the P4-PSA spec \emph{can and should be used} to drive asynchronous environmental control.
We explain relevant NFP architectural details later in \cref{sec:netronome-platform-fundamentals}.
\approachshort{} communicates with the packet pipeline of a P4 dataplane via \texttt{extern} plugins using \inring{} (state, configuration) and \outring{} (action) messages (\cref{fig:netro-arch}).
Internally, \approachshort{} either has all its cores act independently (\cref{fig:single-and-parallel:single}) or cooperate to solve each task (\cref{fig:single-and-parallel:parallel})---with different latency-throughput benefits (\cref{sec:action-and-update-computation}).
We open-source our firmware and control programs for the benefit of the community.\footnote{\textanon{\url{https://github.com/FelixMcFelix/pdp-rl-paper}}{Link redacted for double-blind submission.}}
%Moving beyond the Netronome platform, we describe how our architecture may be adapted and improved on by bespoke hardware or FPGA-based deployment.

@@

\subsection{Action and Update Computation}\label{sec:action-and-update-computation}
\approachshort{} applies the insights of \textcite{DBLP:journals/firai/TravnikMSP18} to minimise action latency; an action is computed, sent out into the environment, and only then is the underlying policy updated.
Using one of the below strategies chosen at compile time, a state vector is tile coded, converted into action probabilities, and an action is chosen.
This is then written out to the environment as in \cref{sec:agent-environment-communication}.
If online learning is enabled, \approachshort{} then checks an internal hashmap for a previous state-action pair matching the current instruction source, and if found then the policy is updated.
Updates are computed using \emph{single-step semi-gradient Sarsa}~\cite[pp. \numrange{217}{221}]{RL2E}, though modification to support other single-step methods would be trivial.
The new state-action pair is then written into storage.
\approachshort{} can be configured to automatically select values of the input state vector as keys for state and reward storage.
Two firmware models govern how these tasks (action selection, policy updates) are carried out:
\begin{description}
	\item[\Indfw{} (\cref{fig:single-and-parallel:single})] Separate threads listen for new states, and each works sequentially. Computing an action list requires a \emph{read lock} on the policy. If an update occurs, the core requests a \emph{write lock} before updating, limiting online throughput. \emph{Tile lists} are stored for update computation.
	\item[\Coopfw{} (\cref{fig:single-and-parallel:parallel,alg:parsa})] Threads cooperate to process state vectors, minimising latency. \emph{Minion} threads have a fixed list of work items, while a \emph{controller} thread sends compute/update commands before awaiting worker completion. Work items are disjoint, requiring no policy locks. \emph{State vectors} are stored for update computation.
\end{description}
Each offers a different point of optimisation; if updates are disabled, then the \indfw{} model can maximise throughput, while the \coopfw{} model is designed to minimise decision latency and needs no locks to update the policy (increasing \emph{online learning} throughput).
These correspond to only executing a trained policy and actively (re-)training a policy, respectively.
%ParSa is described here at a high level, as an algorithm suited for \emph{any multiprocessor environment}.
%We omit the simple logic for $\epsilon$-greedy action selection (which we implement), and note that modification to other single-step algorithms such as \emph{Q-learning} would be trivial.
%We detail our communication primitives in \cref{sec:intra-agent-communication}, and our work scheduling strategy in \cref{sec:work-allocation}, eliding the details of tile coding as they are well-understood.
%In general it is also possible for the \emph{Ctl} task to act as an additional \emph{Minion} in its parallel sections; we were limited here by code store requirements on the NFP.
Latency and throughput have different effects on RL agents according to their design and target problem.
Higher RL throughput is a necessity for per-flow or per-packet applications, which can require high decision-per-second rates even after combining state measurements received from the environment, such as flow control in DDoS prevention.
Equally, lower latency affords an agent finer-grained control and learning of a problem, being able to react sooner to new information (e.g., device state in a routing optimisation problem).

@@

\section{Evaluation}\label{sec:evaluation}
We investigate the performance of \approachshort{} compared to classical RL techniques executed on commodity host machines, with \Coopfw{} offering a \SIrange{15}{21}{$\times$} speedup in median--\num{99.99}\nthscript{th} state-action latency and \SI{9.9}{$\times$} greater online learning throughput.
Crucially, in-NIC execution offers tight tail latency bounds compared to host-based approaches.
We report on how \approachshort{} scales as additional device resources are added, noting that both in-NIC designs outperform commodity hosts using just one core in latency and online throughput.
Furthermore, \Indfw{} provides higher per-core offline throughput than host-based approaches, even though our measured hosts exhibit higher clock speeds.
Finally, we show that \approachshort{} has minimal impact on dataplane cross-traffic carried by its parent device.

@@

\subsection{Experimental Setup}\label{sec:experimental-setup}
Testing machines were as follows, with \SI{32}{\gibi\byte} RAM:
\begin{description}
	\item[\emph{MidServer}] Intel Xeon Bronze 3204 (\qtyproduct[product-units=single]{6 x 1.9}{\giga\hertz}),
	\item[\emph{HighServer}] Intel Xeon Silver 4208 (\qtyproduct[product-units=single]{8 x 2.1}{\giga\hertz}),
	\item[\emph{Collector}] Intel Core i7-6700K (\qtyproduct[product-units=single]{4 x 4.2}{\giga\hertz}).
\end{description}
\approachshort{} was evaluated on server blades (\emph{Mid/HighServer}), each hosting a single Netronome Agilio LX \numproduct{1 x 40}GbE SmartNIC (NFP-6480, \SI{1.2}{\giga\hertz}).
These servers ran Ubuntu 18.04.5 LTS (4.15.0-140-generic).
We additionally use a more powerful consumer-grade machine (\emph{Collector}) for estimating host performance when offloaded to a network function, running Ubuntu 18.04.4 LTS (4.15.0-96-generic).
Control programs were built using rustc 1.52.1.
We run \approachshort{} on a \num{4}-ME island of the NFP-6480 (\num{32} contexts) using \SIlist{32;16;8}{\bit} quantised arithmetic.
This is the largest cluster of cores which is not in use by a P4 pipeline.
All \approachshort{} timing measurements were repeated over \num{10000} state packets (preceded by \num{1000} warmup packets), retrieving item processing times over PCIe via the controller machine, from which throughput was derived.
Host throughput and latency measures were observed over \num{10} trials of \SI{10}{\second} (with \SI{5}{\second} warmup/cooldown times).
We differ this from the NFP as hosts need to run numpy-based agents in parallel via separate processes; this also allows us to investigate the effects of oversubscription on host latencies.

@@

\subsection{Results and Discussion}\label{sec:results}
\fakepara{Raw inference and learning performance}
\Cref{tab:lats} shows how \approachshort{} compares in latency with a numpy-based RL policy.\footnote{For brevity, we omit integer numpy results---against a float implementation, median action latencies are \SI{14.6}{\percent} worse, with \SI{7.9}{\percent} longer update times.}
We show that when using 4 MEs of the NFP-6480, \Coopfw{} achieves sub-\SI{35}{\micro\second} median latency, with \nth{99} and \num{99.99}\nthscript{th} percentile latencies less than \SI{1}{\micro\second} worse (\SIlist{15;21}{$\times$} speedups over a \emph{Collector} host).
Importantly, \Indfw{} achieves lower median state-action latencies (\SI{2.79}{$\times$}) \emph{and} update times (\SI{2.63}{$\times$}) than a dedicated \emph{Collector} host \emph{while requiring only a single core or dedicated functional unit}.
Crucially, in-NIC execution gives far tighter bounds on tail latency compared to host offloading.
\num{99.99}\nthscript{th} percentile state-action latencies exceed the median by \SIlist{0.58;0.66}{\percent} for \Indfw{} and \Coopfw{}, while host-based tail latencies are at least \SI{40.53}{\percent} greater even when the parallel worker count is at or below the physical core count.
We show the cumulative distributions of these in detail (\cref{fig:lat-cumul}), noting how just one additional CPU-intensive task---potentially automated system updates, scans, or another vNF/traffic processing task---impacts tail latencies further (\emph{Float(Over)}).

@@

\begin{figure}[t]
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core}
	}
	\caption{
		\Coopfw{}'s online learning performance improves with additional cores, on max size tasks. This requires \num{8} workers to offer greater online throughput than single-threaded in-NIC RL. Sharper performance increases occur when a new physical core is added (\numrange{7}{8}) or the scheduler works around a bottleneck (\numrange{13}{14}).\label{fig:vary-core}}
\end{figure}

>>>>

% !TeX document-id = {5530719d-34df-4dd8-b4b5-e6ed092c2b36}
% !TeX program = pdflatex
% !BIB program = biber
\documentclass[
%sigconf,natbib=false
%,anonymous=true
conference
,10pt
]{IEEEtran}

%\usepackage[subtle,bibnotes,charwidths,mathdisplays,indent,lists]{savetrees}
%\usepackage[subtle, indent, lists]{savetrees}

@@

@@

%\usepackage[title]{appendix}

@@

\usepackage[font=footnotesize]{subcaption}

@@

\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\newcommand{\crefrangeconjunction}{--}
\crefname{table}{table}{tables}

@@

\newcommand{\subparagraph}{}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.5ex}{0.7ex}
\titlespacing*{\subsection}{0pt}{1.5ex}{0.7ex}
\titlespacing*{\paragraph}{0pt}{1.5ex}{0.7ex}
\titlespacing*{\subsubsection}{0pt}{1.5ex}{0.7ex}

@@

% Copyright

@@

%\newcounter{insightc}
%\newenvironment{insight}
%	{
%		\begin{tipblock}\refstepcounter{insightc}\textbf{Insight \theinsightc:}\em
%	}
%	{
%		\end{tipblock}
%	}

\usepackage[hang,flushmargin]{footmisc}

@@

%	\author{Paper \# XXX, XXX pages}
	 \author{Kyle A. Simpson, Dimitrios P. Pezaros,\\University of Glasgow, Glasgow, Scotland}

%\setlength{\abovecaptionskip}{5pt}
%\setlength{\belowcaptionskip}{0pt}
%
%\setlength{\textfloatsep}{3.0pt plus 0.0pt minus 3.0pt}
%\setlength{\floatsep}{3.0pt plus 0.0pt minus 2.0pt}
%
%\setlength{\dbltextfloatsep}{3.0pt plus 0.0pt minus 3.0pt}
%\setlength{\dblfloatsep}{3.0pt plus 0.0pt minus 2.0pt}

@@

We present \emph{\approachshort{}}---\approach{}---the first work to bring \emph{online reinforcement learning} to the dataplane.
\approachshort{} makes online learning possible in constrained SmartNIC hardware by returning to classical RL techniques which simplify update logic and enable parallel processing.
Compared to offloading, we achieve a \SI{21}{\texttimes} reduction in \num{99.99}\nthscript{th} tail inference times to \SI{34}{\micro\second}, and \SI{9.9}{\texttimes} improvement in online throughput for real-world policy designs.
In-NIC execution eliminates PCIe transfers, and our asynchronous compute model ensures minimal impact on traffic carried by a co-hosted P4 dataplane.
\approachshort{}'s design scales with additional resources at compile-time to improve upon both decision latency and throughput, improving on offloading, and are quickly reconfigurable at runtime compared to reinstalling device firmware.

@@

@@

%\setlength{\aweboxleftmargin}{0.12\linewidth}
%\setlength{\aweboxcontentwidth}{1.97\linewidth}

@@

%The question we investigate is: can online RL be brought to the dataplane by returning to these computationally simpler methods, to act on locally extracted state?
%Can it be made more efficient \emph{by dataplane hardware}?
%Through this work, \approachshort{}, we can comfortably answer ``yes'' on both counts.
%Through \approachshort{}, w
We show that online RL can be brought to the dataplane by such methods, where it can act on locally extracted state and is made \emph{more efficient by dataplane hardware}.
In particular, we exploit how SmartNIC devices often expose general-purpose compute to provide path-adjacent, on-chip RL in the dataplane (\cref{fig:netro-arch}).
As many of these devices have engineering and development histories which predate P4, general compute beyond P4's limits~\parencite{p4-psa} is surprisingly common.
By executing on spare compute units, we prevent packet stalling and offer quick runtime reconfigurability.
%We look backwards to classical RL techniques (tile-coded policies, single-step RL algorithms), alongside the adaptations introduced by existing in-network ML (such as quantisation).
%Such adaptations unlock further improvements in parallel inference and update processing.
This paper contributes:
\begin{itemize}
	\item An analysis of why in-NIC RL is needed and best-placed to interact with the network, made feasible by classical RL methods and quantisation (\cref{sec:motivation}),
	\item \emph{\approachshort{}}: a general-purpose in-NIC RL agent which scales with allocated device resources to meet latency or throughput demands of network traffic analysis (\cref{sec:design}),
	\item \emph{ParSa}, a wait-free, parallel, online RL algorithm to accelerate tile coded policy inference and updates (\cref{alg:parsa}),
	\item In-depth evaluation of how \approachshort{} affects carried dataplane traffic, performs under different policy sizes (simple/complex state), and improves on explicit offloading with a \SI{15}{\texttimes} latency reduction compared to commodity hardware (\SI{21}{\texttimes} for 99.99\nthscript{th} tail latencies) and an order of magnitude improvement in online throughput (\cref{sec:evaluation}).
	\item A description of how \approachshort{} would integrate with state-of-the-art PDP applications to perform fully in-NIC, fast, automated DDoS mitigation (\cref{sec:potential-integrations}).
\end{itemize}

@@

%\begin{figure}
%	\begin{subfigure}{0.45\linewidth}
%		\centering
%		\resizebox{0.8\linewidth}{!}{
%			\centering
%			\begin{tikzpicture}
%				\node[circle, draw] (state) {$S$};
%				\node[circle, draw] (state') at ($(state) + (0, -2)$) {$S'$};
%				
%				\node (agent) at ($(2, 0) + (state)$) {Agent};
%				\node[below of=agent] (action) {Action $A$};
%				
%				\node[right of=state'] {$+ R$};
%				
%				\draw[->] (state) -- (state') node[midway, right] {$A$};
%				
%				\draw[dotted, ->, bend left = 30] (state) -- (agent);
%				\draw[->] (agent) -- (action);
%				\draw[dotted, ->] (action) -- (state);
%		\end{tikzpicture}}
%		\caption{Theory: state measurement, action computation, and learning are zero-cost.}
%	\end{subfigure}
%	\hspace{0.05\linewidth}
%	\begin{subfigure}{0.45\linewidth}
%		\centering
%		\resizebox{0.6\linewidth}{!}{
%			\centering
%			\begin{tikzpicture}
%				\node[circle, draw] (state) {$S$};
%				\node[circle, draw] (state') at ($(state) + (0, -1.5)$) {$S'$};
%				\node[circle, draw] (state'') at ($(state') + (0, -1.5)$) {$S''$};
%				
%				\node (agent) at ($(2, 0) + (state)$) {Agent};
%				\node[below of=agent] (action) {Action $A$};
%				
%				\node[right of=state''] {$+ R$};
%				
%				\draw[->] (state) -- (state') node[midway, right] {$\phi$}; %$\varnothing$};
%				\draw[->] (state') -- (state'') node[midway, right] {$A$};
%				
%				\draw[dotted, ->, bend left = 30] (state) -- (agent) node[midway, above] {$t_1$};
%				\draw[->] (agent) -- (action) node[midway, right] {$t_2$};
%				\draw[dotted, ->] (action) -- (state') node[midway, below] {$t_3$};
%		\end{tikzpicture}}
%		\caption{Reality: policy costs lead to \emph{state drift}---over a delay $t_1+t_2+t_3$, state $S$ becomes $S'$.}
%	\end{subfigure}
%	\caption{Asynchronous RL delays and state slippage.\label{fig:state-slip}}
%\end{figure}

@@

%\begin{figure*}
%	\centering
%	\begin{subfigure}{0.47\linewidth}
%		\centering
%		\includegraphics[keepaspectratio, width=0.78\linewidth]{figures/ind}
%		\caption{\Indfw{} (offline throughput-optimal). \emph{Workers} independently pull commands from (and push actions to) the environment, locking policy access for updates.\label{fig:single-and-parallel:single}}
%	\end{subfigure}
%	\hspace{0.04\linewidth}
%	\begin{subfigure}{0.47\linewidth}
%		\centering
%		\includegraphics[keepaspectratio, width=0.8\linewidth]{figures/coop}
%		\caption{\Coopfw{} (online-optimal). A single \emph{controller} delegates RL computation and updates to many \emph{minion} threads, who operate on independent subtasks.\label{fig:single-and-parallel:parallel}}
%	\end{subfigure}
%	\caption{\approachshort{}'s compute strategies scale to fit device capacity according to either latency or throughput needs.\label{fig:single-and-parallel}}
%\end{figure*}

\begin{figure}
	\centering
	\includegraphics[keepaspectratio, width=0.8\linewidth]{figures/coop}
	\caption{\approachshort{}'s compute strategies scale to fit device capacity according to either latency or throughput needs---\Coopfw{} shows this optimally online. A single \emph{controller} delegates RL computation and updates to many \emph{minion} threads, who operate on independent subtasks.\label{fig:single-and-parallel:parallel}}
\end{figure}

\Cref{fig:netro-arch,fig:single-and-parallel} outline our design on Netronome SmartNIC hardware in pursuit of this goal: unused device resources beyond the P4-PSA spec \emph{can and should be used} to drive asynchronous environmental control.
We explain relevant NFP architectural details later in \cref{sec:netronome-platform-fundamentals}.
\approachshort{} communicates with the packet pipeline of a P4 dataplane via \texttt{extern} plugins using \inring{} (state, configuration) and \outring{} (action) messages (\cref{fig:netro-arch}).
Internally, \approachshort{} either has all its cores act independently (\cref{fig:single-and-parallel:single}) or cooperate to solve each task (\cref{fig:single-and-parallel:parallel})---with different latency-throughput benefits (\cref{sec:action-and-update-computation}).
%We open-source our firmware and control programs for the benefit of the community.\footnote{\textanon{\url{https://github.com/FelixMcFelix/pdp-rl-paper}}{Link redacted for double-blind submission.}}
We open-source our firmware and control programs for the benefit of the community.\footnote{Link redacted during submission.}
%Moving beyond the Netronome platform, we describe how our architecture may be adapted and improved on by bespoke hardware or FPGA-based deployment.

@@

\subsection{Action and Update Computation}\label{sec:action-and-update-computation}
\approachshort{} applies the insights of \textcite{DBLP:journals/firai/TravnikMSP18} to minimise action latency; an action is computed, sent out into the environment, and only then is the underlying policy updated.
Using one of the below strategies chosen at compile time, a state vector is tile coded, converted into action probabilities, and an action is chosen.
This is then written out to the environment as in \cref{sec:agent-environment-communication}.
If online learning is enabled, \approachshort{} then checks an internal hashmap for a previous state-action pair matching the current instruction source, and if found then the policy is updated.
Updates are computed using \emph{single-step semi-gradient Sarsa}~\cite[pp. \numrange{217}{221}]{RL2E}, though modification to support other single-step methods would be trivial.
The new state-action pair is then written into storage.
\approachshort{} can be configured to automatically select values of the input state vector as keys for state and reward storage.
Two firmware models govern how these tasks (action selection, policy updates) are carried out:
\begin{LaTeXdescription}
	\item[\Indfw{} (\cref{fig:single-and-parallel:single})] Separate threads listen for new states, and each works sequentially. Computing an action list requires a \emph{read lock} on the policy. If an update occurs, the core requests a \emph{write lock} before updating, limiting online throughput. \emph{Tile lists} are stored for update computation.
	\item[\Coopfw{} (\cref{fig:single-and-parallel:parallel,alg:parsa})] Threads cooperate to process state vectors, minimising latency. \emph{Minion} threads have a fixed list of work items, while a \emph{controller} thread sends compute/update commands before awaiting worker completion. Work items are disjoint, requiring no policy locks. \emph{State vectors} are stored for update computation.
\end{LaTeXdescription}
Each offers a different point of optimisation; if updates are disabled, then the \indfw{} model can maximise throughput, while the \coopfw{} model is designed to minimise decision latency and needs no locks to update the policy (increasing \emph{online learning} throughput).
These correspond to only executing a trained policy and actively (re-)training a policy, respectively.
%ParSa is described here at a high level, as an algorithm suited for \emph{any multiprocessor environment}.
%We omit the simple logic for $\epsilon$-greedy action selection (which we implement), and note that modification to other single-step algorithms such as \emph{Q-learning} would be trivial.
%We detail our communication primitives in \cref{sec:intra-agent-communication}, and our work scheduling strategy in \cref{sec:work-allocation}, eliding the details of tile coding as they are well-understood.
%In general it is also possible for the \emph{Ctl} task to act as an additional \emph{Minion} in its parallel sections; we were limited here by code store requirements on the NFP.
Latency and throughput have different effects on RL agents according to their design and target problem.
Higher RL throughput is a necessity for per-flow or per-packet applications, which can require high decision-per-second rates even after combining state measurements received from the environment, such as flow control in DDoS prevention.
Equally, lower latency affords an agent finer-grained control and learning of a problem, being able to react sooner to new information (e.g., device state in a routing optimisation problem).

@@

\section{Evaluation}\label{sec:evaluation}
We investigate the performance of \approachshort{} compared to classical RL techniques executed on commodity host machines, with \Coopfw{} offering a \SIrange{15}{21}{\texttimes} speedup in median--\num{99.99}\nthscript{th} state-action latency and \SI{9.9}{\texttimes} greater online learning throughput.
Crucially, in-NIC execution offers tight tail latency bounds compared to host-based approaches.
We report on how \approachshort{} scales as additional device resources are added, noting that both in-NIC designs outperform commodity hosts using just one core in latency and online throughput.
Furthermore, \Indfw{} provides higher per-core offline throughput than host-based approaches, even though our measured hosts exhibit higher clock speeds.
Finally, we show that \approachshort{} has minimal impact on dataplane cross-traffic carried by its parent device.

@@

\subsection{Experimental Setup}\label{sec:experimental-setup}
Testing machines were as follows, with \SI{32}{\gibi\byte} RAM:
\begin{LaTeXdescription}
	\item[\emph{MidServer}] Intel Xeon Bronze 3204 (\qtyproduct[product-units=single]{6 x 1.9}{\giga\hertz}),
	\item[\emph{HighServer}] Intel Xeon Silver 4208 (\qtyproduct[product-units=single]{8 x 2.1}{\giga\hertz}),
	\item[\emph{Collector}] Intel Core i7-6700K (\qtyproduct[product-units=single]{4 x 4.2}{\giga\hertz}).
\end{LaTeXdescription}
\approachshort{} was evaluated on server blades (\emph{Mid/HighServer}), each hosting a single Netronome Agilio LX \numproduct{1 x 40}GbE SmartNIC (NFP-6480, \SI{1.2}{\giga\hertz}).
These servers ran Ubuntu 18.04.5 LTS (4.15.0-140-generic).
We additionally use a more powerful consumer-grade machine (\emph{Collector}) for estimating host performance when offloaded to a network function, running Ubuntu 18.04.4 LTS (4.15.0-96-generic).
Control programs were built using rustc 1.52.1.
We run \approachshort{} on a \num{4}-ME island of the NFP-6480 (\num{32} contexts) using \SIlist{32;16;8}{\bit} quantised arithmetic.
This is the largest cluster of cores which is not in use by a P4 pipeline.
All \approachshort{} timing measurements were repeated over \num{10000} state packets (preceded by \num{1000} warmup packets), retrieving item processing times over PCIe via the controller machine, from which throughput was derived.
Host throughput and latency measures were observed over \num{10} trials of \SI{10}{\second} (with \SI{5}{\second} warmup/cooldown times).
We differ this from the NFP as hosts need to run numpy-based agents in parallel via separate processes; this also allows us to investigate the effects of oversubscription on host latencies.

@@

\subsection{Results and Discussion}\label{sec:results}
\fakepara{Raw inference and learning performance}
\Cref{tab:lats} shows how \approachshort{} compares in latency with a numpy-based RL policy.\footnote{For brevity, we omit integer numpy results---against a float implementation, median action latencies are \SI{14.6}{\percent} worse, with \SI{7.9}{\percent} longer update times.}
We show that when using 4 MEs of the NFP-6480, \Coopfw{} achieves sub-\SI{35}{\micro\second} median latency, with \nth{99} and \num{99.99}\nthscript{th} percentile latencies less than \SI{1}{\micro\second} worse (\SIlist{15;21}{\texttimes} speedups over a \emph{Collector} host).
Importantly, \Indfw{} achieves lower median state-action latencies (\SI{2.79}{\texttimes}) \emph{and} update times (\SI{2.63}{\texttimes}) than a dedicated \emph{Collector} host \emph{while requiring only a single core or dedicated functional unit}.
Crucially, in-NIC execution gives far tighter bounds on tail latency compared to host offloading.
\num{99.99}\nthscript{th} percentile state-action latencies exceed the median by \SIlist{0.58;0.66}{\percent} for \Indfw{} and \Coopfw{}, while host-based tail latencies are at least \SI{40.53}{\percent} greater even when the parallel worker count is at or below the physical core count.
We show the cumulative distributions of these in detail (\cref{fig:lat-cumul}), noting how just one additional CPU-intensive task---potentially automated system updates, scans, or another vNF/traffic processing task---impacts tail latencies further (\emph{Float(Over)}).

@@

\begin{figure}[t]
	\centering
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core}
	}
	\caption{
		\Coopfw{}'s online learning performance improves with additional cores, on max size tasks. This requires \num{8} workers to offer greater online throughput than single-threaded in-NIC RL. Sharper performance increases occur when a new physical core is added (\numrange{7}{8}) or the scheduler works around a bottleneck (\numrange{13}{14}).\label{fig:vary-core}}
\end{figure}
